D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-26 ootavacuous.tex
January 26th reorganization, part 1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	We present informal arguments that these constraints prevent
	OOTA cycles containing only volatile atomics from occurring
	in UB-free C++ programs running on correctly constructed C++
	implementations that are restricted to single-threaded analysis
	in our current universe.
	We further conjecture that, given current optimizations, these
	constraints also prevent OOTA cycles involving non-volatile
	atomics.
\end{abstract}

@@

However, it is important to distinguish OOTA cycles from the OOTA-like
behavior that can be caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ implementation\footnote{
	This paper will use ``C++ implementation'' both excluding the
	CPU, as here, and including the CPU.
	Explicit disambiguation is added when necessary.}
and the CPU are within their rights to reorder lines~12 and~13, which
can result in all of \co{x}, \co{y}, \co{r1}, and \co{r2} having the
value 42 as follows:

@@

In contrast,
Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
shows a simple example where an OOTA cycle might result in all of \co{x},
\co{y}, \co{r1}, and \co{r2} have final values of 42, despite the fact
that there is nothing in the code or in the initial values that would
support such an outcome:

@@

Roughly speaking, \emph{there is semantic dependency from a given load
to a given store when a change in the value loaded that can change the
value stored.}

@@

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   unsigned long r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   unsigned long r2 = y.load(memory_order_relaxed);
13   x.store(42, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple Reordering}
\label{lst:Simple Reordering}
\end{listing}

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   unsigned long r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   unsigned long r2 = y.load(memory_order_relaxed);
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple OOTA}
\label{lst:Simple OOTA}
\end{listing}

@@

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes his examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

All this work focused on either identifying OOTA or on how C++
implementations could avoid it.
None of this work applied real-world temporal constraints to the OOTA
problem.
Which might explain why no known real-world C++ implementation results
in OOTA executions.

@@

This paper will use the \co{herd7} tool to analyze fragments of C++ code.
This tool carries out the moral equivalent of full state-space searches
of concurrent code fragments, and further has some understanding of OOTA.

@@

The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first process, \co{P0()}, and lines~15-21
define the second process, \co{P1()}.
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each process may access, in this case, \co{x}
and \co{y}.
The body of each process specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

@@

Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
of the \co{herd7} tool.
Lines~3 and~4 show the possible states, with line~4 showing the
counterintuitive outcome where both processes load the value 42.
Normally these states would include only those variables mentioned
in the \co{exists} clause, but because of the \co{locations} clause
on line~23, the values of \co{x} and \co{y} is also included, which
can be helpful when debugging.
As we will see, this outcome is not possible in real-world implementations
due to physical constraints of which \co{herd7} is (by design) unaware.
Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
indicating that some executions satisfy the \co{exists} clause and
others do not.

@@

This is a fine definition, and is consistent with the words in the C++
standard, but has an intra-thread problem exemplified by
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
This is an elaboration of
Listing~\ref{lst:Simple OOTA}
that adds \co{z} and adds lines~15 and~16 to \co{thread2()}.
The problem is that the C++ implementation is within its rights to
note that line~16 might well execute immediately after line~15, giving
threads no chance to modify \co{z}.
Such an implementation might therefore behave as if the source code
had instead been as shown in
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
with line~16 optimized away in favor of line~17 storing the value
in \co{r2} that was loaded from \co{y}.
And given that the load from \co{z} no longer exists, it cannot possibly
act as a temporal constraint.

@@

This wording presumably includes the initial value of M as one of the
possible side effects.
This means that an implementation might determine that a given atomic
object always has a given value, and such an implementation might
skip the load in favor of that value.\footnote{
	There is some work starting to forbid the implementation
	from using this assumption, for example, by requiring that
	the implementation assume that there is at least one thread
	that it does not know about.}

Please note that this does not apply to volatile atomic objects,
for which the implementation is forbidden from assuming that the
value loaded has any relationship whatsoever to any initialization
or any values stored.

@@

\begin{quote}
	\co{x = y * z;}
\end{quote}

@@

\begin{quote}
	\co{x = y + z;}
\end{quote}

There are semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation knows that the value of \co{y} is always
the negative of \co{z}, there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were
instead the constant zero, again, at least until the next assignment
to \co{x}.

@@

\begin{quote}
	\co{r1 = x;} \\
	\co{if (r1 > 0)} \\
	\co{\ \ \ \ y = r1;} \\
	\co{else} \\
	\co{\ \ \ \ z = r1;}
\end{quote}

@@

@@

\begin{quote}
	\co{r1 = x;} \\
	\co{if (r1 > 0)} \\
	\co{\ \ \ \ y = 42;} \\
	\co{else} \\
	\co{\ \ \ \ y = 42;}
\end{quote}

Because the stores executed on each branch of that "if" statement
store identical values to identical addresses, they are for all intents
and purposes the same store.
Therefore, there is no semantic dependency~\cite{Howells2009membartxt}.

@@

\begin{quote}
	\co{r1 = x;} \\
	\co{if (r1 > 42)} \\
	\co{\ \ \ \ r2 = y;} \\
	\co{else} \\
	\co{\ \ \ \ r2 = y;} \\
	\co{z = r2;}
\end{quote}

@@

\subsection{@@@ This marks the end of the reworked material}

@@

The earliest work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision process for what does and does not
constitute a causal cycle.
P0442R0 defines an OOTA cycle as a fixed-point computation that is
destroyed by perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.
Still others argue (perhaps correctly) that it is impossible to provide
a general definition of ``OOTA cycle'', but this simply raises
the question of whether a \emph{useful} definition can be formulated.

@@

\begin{quote}
	\co{r1 = x;} \\
	\co{// r1 = 0;} \\
	\co{z = y * r1;} \\
	\co{// z = y / r1;}
\end{quote}

@@

\begin{quote}
	\co{r1 = x;} \\
	\co{z = y * r1;} \\
	\co{z = 42;}
\end{quote}

@@

\begin{quote}
	\co{int foo(int a, int b)} \\
	\co{\{} \\
	\co{\ \ \ \ return a / b;} \\
	\co{\}} \\
	\\
	\co{y = foo(++i, ++i);} \\
	\co{x = y * z;}
\end{quote}

@@

\begin{quote}
	\co{x = y + z;}
\end{quote}

@@

\begin{quote}
	\co{x = 2 * y;}
\end{quote}

@@

\begin{quote}
	\co{x = 2 * z;}
\end{quote}

@@

\begin{quote}
	\co{x = y * input_int();}
\end{quote}

@@

\begin{quote}
	\co{x = horribly_complex(y, z);}
\end{quote}

@@

\begin{quote}
	\co{x = y & 0x1 ? do_odd(y) : do_even(y);}
\end{quote}

@@

\begin{quote}
	\co{x = collatz_cycle_min(y);}
\end{quote}

@@

\begin{quote}
	\co{void foo(int x, char c)} \\
	\co{\{} \\
	\co{\ \ \ \ return x * (c >= 0);} \\
	\co{\}}
\end{quote}

@@

@@

This optimization process may, roughly speaking, be divided into two
activities, analysis of the code and classification of dependencies.
Two sections on these topics are followed by sections on optimizations
and on consequences.

@@

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe''), and loads from objects for which the value
returned by a given load was overwritten by some other thread
(``fre'').\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for more
	information on these memory-reference interprocess communications
	links.}

@@

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} after the other two processes, and C++ is not
constrained to do this.

@@

Both processes have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} before the other two processes, and C++ is not
constrained to do this.

@@

All processes have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

All processes's loads' return values affect the value stored, so all
processes have semantic dependencies between their respective relaxed
atomic loads and relaxed atomic stores.
Therefore, an OOTA cycle cannot form in a correct real-world C++
implementation.

@@

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
shows a different way of expanding
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two processes to three processes.
The \co{herd7} tool finds OOTA values, as expected.

@@

All processes have straightforward semantic dependencies from their
relaxed atomic loads to their relaxed atomic stores, which prevents this
OOTA cycle on correct real-world C++ implementations.

@@

Although the examples presented in
Section~\ref{sec:OOTA: rf vs. rfe}
demonstrate that rf can be atemporal, they demonstrate neither
OOTA cycles nor OOTA-like behavior in full C++ implementations.
This appendix provides more involved examples that can be optimized
so as to demonstrate OOTA-like behavior.

@@

\begin{quote}
	\co{z = w * x + w * y + w;}
\end{quote}

@@

Listing~\ref{lst:Example Code for Evaluation of sdep}
shows a \co{herd7} process \co{P0()}.
Is there a semantic dependency between the load from \co{x} on
line~3 to the store to \co{y} on line~9?

@@

One way to answer this question is to create a C program
(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
that runs the \co{P0()} process's code twice with arbitrary inputs
(the loop spanning lines~6-17) and asserts that the results of
each iteration be equal (line~18).

@@

\begin{quote}
	\scriptsize
	\co{cbmc -DCBMC cbmc/sdep-quadratic.c}
\end{quote}

@@

\begin{quote}
	\scriptsize
	\co{cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c}
\end{quote}

@@

\begin{quote}
	\scriptsize
	\co{cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c}
\end{quote}

@@

Both the ARM and PowerPC memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of any such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering.
% @@@ Cite Gristenthwaite

@@

However, some code has ways of handling even uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one process, and might react by
killing that process.
Some user applications might be note that the affected data affected only
a particular computation, and might react by restarting that computation.
Both the kernel and the user application would need to continue execution,
and might thus desire an exact exception.

@@

>>>>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We further conjecture that, given current optimizations, these
	constraints also prevent OOTA cycles involving non-volatile
	atomics.
\end{abstract}

@@

However, it is important to distinguish OOTA cycles from the OOTA-like
behavior that can be caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder
lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 as follows:

@@

In contrast,
Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
shows a simple example where an OOTA cycle might result in all of \co{x},
\co{y}, \co{r1}, and \co{r2} have final values of 42, despite the fact
that there is nothing in the initial values or the executable code that
would support such an outcome:

@@

Roughly speaking and all other things being equal, \emph{there is semantic
dependency from a given load to a given store when a change in the value
loaded that can change the value stored.}

@@

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = y.load(memory_order_relaxed);
13   x.store(42, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple Reordering}
\label{lst:Simple Reordering}
\end{listing}

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = y.load(memory_order_relaxed);
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple OOTA}
\label{lst:Simple OOTA}
\end{listing}

@@

Goldblatt looked at interactions between OOTA cycles and
undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes his examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

All this work focused on either identifying OOTA or on how C++
implementations could avoid it.
None of this work applied real-world hardware ordering to the problem
of avoiding OOTA cycles.
Which might explain why no known real-world C++ implementation results
in OOTA executions.

@@

This paper will use the \co{herd7} tool to analyze fragments of C++ code.
This tool carries out the moral equivalent of full state-space searches
of concurrent code fragments, and further will in some cases identify
OOTA cycles.

@@

The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first thread, \co{P0()}, and lines~15-21
define the second thread, \co{P1()}.\footnote{
	The ``P'' stands for ``process'', which is \co{herd7}'s name
	for ``thread''.}
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each thread may access, in this case, \co{x}
and \co{y}.
The body of each thread specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

@@

Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
of the \co{herd7} tool.
Lines~3 and~4 show the possible states, with line~4 showing the
counterintuitive outcome where both threads load the value 42.
Normally these states would include only those variables mentioned
in the \co{exists} clause, but because of the \co{locations} clause
on line~23, the values of \co{x} and \co{y} is also included, which
can be helpful when debugging.
% @@@ As we will see, this outcome is not possible in real-world
% @@@ implementations due to physical constraints of which \co{herd7}
% @@@ is (by design) unaware.
Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
indicating that some executions satisfy the \co{exists} clause and
others do not.

@@

This is a fine definition, and is consistent with the words in the C++
standard, but has an intra-thread problem exemplified by the following
code:

\begin{quote}
\scriptsize
\begin{verbatim}
 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
 3   int r3 = atomic_load_explicit(&y, memory_order_relaxed);
 4   atomic_store_explicit(&z, r3, memory_order_relaxed);
\end{verbatim}
\end{quote}

This is an elaboration of \co{thread2()} from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~2 and~3.
The problem is that the C++ implementation is within its rights to
note that line~3 might well execute immediately after line~2, giving
threads no chance to modify \co{y}.
Such an implementation might therefore behave as if the source code
had instead been as follows:

\begin{quote}
\scriptsize
\begin{verbatim}
 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
 3   // int r3 = atomic_load_explicit(&y, memory_order_relaxed);
 4   atomic_store_explicit(&z, r2, memory_order_relaxed);
\end{verbatim}
\end{quote}

Here line~3 has been optimized away in favor of line~4 storing the value
in \co{r2} that was loaded from \co{y} by line~2.
And given that the load from \co{z} no longer exists, it cannot possibly
act as a temporal constraint.

@@

@@

\begin{quote}
\begin{verbatim}
x = y * z;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = y + z;
\end{verbatim}
\end{quote}

There are semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation knows that the value of \co{y} is
always the negative of \co{z} there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were instead
the constant zero, again, at least until the next assignment to \co{x}.
Please note that this leaves aside the question of how the compiler
would know this if \co{y} and \co{z} are atomic objects that cannot be
updated pairwise atomically and are subject to change at any time by
other threads.

@@

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 0)
    y = r1;
else
    z = r1;
\end{verbatim}
\end{quote}

@@

Which raises the question:
What exactly is an execution?

\section{@@@ This marks the end of the reworked material}

\subsection{Difficulties in Rigorously Defining ``Semantic Dependency''}

@@

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}

Because the stores executed on each branch of that "if" statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should take the single-store
viewpoint~\cite{Howells2009membartxt},
the fact remains that these two viewpoints lead to differing conclusions
about whether there is a semantic dependency.

@@

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 42)
    r2 = y;
else
    r2 = y;
z = r2;
\end{verbatim}
\end{quote}

@@

This wording presumably includes the initial value of M as one of the
possible side effects.
This means that an implementation might determine that a given atomic
object always has a given value, and such an implementation might
skip the load in favor of that value.\footnote{
	There is some work starting to forbid the implementation
	from using this assumption, for example, by requiring that
	the implementation assume that there is at least one thread
	that it does not know about.
	\url{https://hackmd.io/@gonzalob/Bk_ufPZ6o/edit}}

Please note that this does not apply to volatile atomic objects,
for which the implementation is prohibited from assuming that the
value loaded has any relationship whatsoever to any initialization
or any values stored.
This prohibition may come as a surprise to those carefully reading the
C++ standard~\cite{ThomasKoeppe2023N4950}
and finding only this wording constraining volatile accessess in
4.1.2p6
\co{[intro.abstract]}:

\begin{quote}
	Accesses through volatile glvalues are evaluated strictly
	according to the rules of the abstract machine.
\end{quote}

Unfortunately, volatile semantics are mostly folklore passed by word
of mouth among some compiler writers and some device-driver developers.
The key point is that volatile accesses must interact with MMIO
registers in such a way as to permit device drivers to be written.
And until such time as the semantics of all MMIO registers of all
devices have been formalized, published, and incorporated into
C++ implementations,\footnote{
	Given that some CPU vendors have started publishing formal
	memory models, this happy day might actually arrive.
	But we strongly recommend against holding your breath while
	waiting for it.}
these implementations cannot asssume that they can predict the value
returned by a volatile load.

@@

The earliest work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision thread for what does and does not
constitute a causal cycle.
P0442R0 defines an OOTA cycle as a fixed-point computation that is
destroyed by perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.
Still others argue (perhaps correctly) that it is impossible to provide
a general definition of ``OOTA cycle'', but this simply raises
the question of whether a \emph{useful} definition can be formulated.

@@

\begin{quote}
\begin{verbatim}
r1 = x;
// r1 = 0;
z = y * r1;
// z = y / r1;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
r1 = x;
z = y * r1;
z = 42;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

y = foo(++i, ++i);
x = y * z;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = y + z;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = 2 * y;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = 2 * z;
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = y * input_int();
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = horribly_complex(y, z);
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = y & 0x1 ? do_odd(y) : do_even(y);
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
x = collatz_cycle_min(y);
\end{verbatim}
\end{quote}

@@

\begin{quote}
\begin{verbatim}
void foo(int x, char c)
{
    return x * (c >= 0);
}
\end{verbatim}
\end{quote}

@@

As noted earlier, this initial work focuses on C++ implementations based
on compilers and computing hardware (including CPUs and GPGPUs).

@@

This optimization thread may, roughly speaking, be divided into two
activities, analysis of the code and classification of dependencies.
Two sections on these topics are followed by sections on optimizations
and on consequences.

@@

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe''), and loads from objects for which the value
returned by a given load was overwritten by some other thread
(``fre'').\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for more
	information on these memory-reference inter-process communications
	links.}

@@

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} after the other two threads, and C++ is not
constrained to do this.

@@

Both threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} before the other two threads, and C++ is not
constrained to do this.

@@

All threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

All thread's loads' return values affect the value stored, so all
threads have semantic dependencies between their respective relaxed
atomic loads and relaxed atomic stores.
Therefore, an OOTA cycle cannot form in a correct real-world C++
implementation.

@@

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
shows a different way of expanding
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two threads to three threads.
The \co{herd7} tool finds OOTA values, as expected.

@@

All threads have straightforward semantic dependencies from their
relaxed atomic loads to their relaxed atomic stores, which prevents this
OOTA cycle on correct real-world C++ implementations.

@@

Although the examples presented in
Section~\ref{sec:OOTA: rf vs. rfe}
(and shown as litmus tests in
Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
demonstrate that rf can be atemporal, they demonstrate neither
OOTA cycles nor OOTA-like behavior in full C++ implementations.
This appendix provides more involved examples that can be optimized
so as to demonstrate OOTA-like behavior.

@@

\begin{quote}
\begin{verbatim}
z = w * x + w * y + w;
\end{verbatim}
\end{quote}

@@

Listing~\ref{lst:Example Code for Evaluation of sdep}
shows a \co{herd7} thread \co{P0()}.
Is there a semantic dependency between the load from \co{x} on
line~3 to the store to \co{y} on line~9?

@@

One way to answer this question is to create a C program
(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
that runs the \co{P0()} thread's code twice with arbitrary inputs
(the loop spanning lines~6-17) and asserts that the results of
each iteration be equal (line~18).

@@

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC cbmc/sdep-quadratic.c
\end{verbatim}
\end{quote}

@@

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
\end{verbatim}
\end{quote}

@@

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
\end{verbatim}
\end{quote}

@@

Both the ARM and PowerPC memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of any such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering.
% @@@ Cite Grisenthwaite

@@

However, some code has ways of handling even uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one thread, and might react by
killing that thread.
Some user applications might be note that the affected data affected only
a particular computation, and might react by restarting that computation.
Both the kernel and the user application would need to continue execution,
and might thus desire an exact exception.

@@

\section{Cutting-Room Floor}
\label{sec:Cutting-Room Floor}

This appendix includes material that was removed but which might later
prove useful.
